---
title: "Practical Machine Learning Course Project Report"
author: "gloriali"
date: "June 20, 2014"
output:
  html_document:
    toc: yes
---

```{r prep, echo = F, message = F}
library(ggplot2)
library(caret)
library(randomForest)
setwd("~/Study/UBC/PracticalMachineLearning")
load("project.Rdata")
```

### Load data 
* Load data.           
* Remove near zero covariates and those with more than 80% missing values since these variables will not provide much power for prediction.       
* Calculate correlations between each remaining feature to the response, `classe`. Use `spearman` rank based correlation because `classe` is a factor.                 
* Plot the two features that have highest correlation with `classe` and color with `classe` to see if we can separate response based on these features.            

```{r data, eval = F}
# load data
training <- read.csv("./data/pml-training.csv", row.names = 1)
testing <- read.csv("./data/pml-testing.csv", row.names = 1)
# remove near zero covariates
nsv <- nearZeroVar(training, saveMetrics = T)
training <- training[, !nsv$nzv]
# remove variables with more than 80% missing values
nav <- sapply(colnames(training), function(x) if(sum(is.na(training[, x])) > 0.8*nrow(training)){return(T)}else{return(F)})
training <- training[, !nav]
# calculate correlations
cor <- abs(sapply(colnames(training[, -ncol(training)]), function(x) cor(as.numeric(training[, x]), as.numeric(training$classe), method = "spearman")))
```
```{r data_plot}
# plot predictors 
summary(cor)
plot(training[, names(which.max(cor))], training[, names(which.max(cor[-which.max(cor)]))], col = training$classe, pch = 19, cex = 0.1, xlab = names(which.max(cor)), ylab = names(which.max(cor[-which.max(cor)])))
```

The training set has __`r nrow(training)`__ samples and __`r ncol(training) - 1`__ potential predictors after filtering.          

There doesn't seem to be any strong predictors that correlates with `classe` well, so linear regression model is probably not suitable in this case. Boosting and random forests algorithms may generate more robust predictions for our data.         

### Boosting model
* Fit model with boosting algorithm and 10-fold cross validation to predict `classe` with all other predictors.    
* Plot accuracy of this model on the scale `[0.9, 1]`.        

```{r boost, eval = F}
set.seed(123)
boostFit <- train(classe ~ ., method = "gbm", data = training, verbose = F, trControl = trainControl(method = "cv", number = 10))
```
```{r boost_plot}
boostFit
plot(boostFit, ylim = c(0.9, 1))
```

The boosting algorithm generated a good model with __accuracy = 0.997__. 

### Random forests model   
* Fit model with random forests algorithm and 10-fold cross validation to predict `classe` with all other predictors.    
* Plot accuracy of the model on the __same scale__ as boosting model.            

```{r rf, eval = F}
set.seed(123)
rfFit <- train(classe ~ ., method = "rf", data = training, importance = T, trControl = trainControl(method = "cv", number = 10))
```
```{r rf_plot}
rfFit
plot(rfFit, ylim = c(0.9, 1))
```
```{r rf_imp, echo = F}
imp <- varImp(rfFit)$importance
imp$max <- apply(imp, 1, max)
imp <- imp[order(imp$max, decreasing = T), ]
```

The random forests algorithm generated a very accurate model with __accuracy close to 1__. Compared to boosting model, this model generally has better performance in terms of accuracy as we see from the plots.                     

### Final model and prediction   
* Comparing model accuracy of the two models generated, random forests and boosting, random forests model has overall better accuracy. So, I'll use this model for prediction.         
* The final random forests model contains 500 trees with 40 variables tried at each split. The five most important predictors in this model are `r rownames(imp)[1:5]`.                 
* Estimated __out of sample error rate__ for the random forests model is __0.04%__ as reported by the final model.            
* Predict the test set and output results for automatic grader.     

```{r prediction, message = F}
# final model
rfFit$finalModel
# prediction
(prediction <- as.character(predict(rfFit, testing)))
```
```{r results, eval = F}
# write prediction files
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./prediction/problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
pml_write_files(prediction)
```

